Reproducibility of SNV-calling in multiple sequencing runs from single tumors

### 2015-09-30 PAPER/paper.tex
3-6/13 rebuttal/revisions

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}. To analyze this data, researchers typically sequence the blood and tumor of a single patient independently and then align them to each other to find somatic variants, any of which could be cancer-causing or -sustaining. TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%. Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}. To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large-scale genomics data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.

@@

Several factors limit the conclusions we may draw from this analysis. First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNV. This metric is potentially problematic for many reasons, including but not limited to: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate and not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; (iii) the SNV calling process may enrich for artifacts such as germ line variants, which are indistinguishable from SNVs in computational analyses and will only be removed from the data if they are recorded in dbSNP; (iv) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}. Although repeatability does not and could not perfectly measure confidence that a putative mutation is a somatic mutation, it does make it more likely. Having an independent measure of confidence in SNV calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.

>>>>

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}. To identify somatic mutations in cancer genomics data, researchers typically independently sequence the blood and tumor of a single patient and align the blood and tumor sequences to each other to identify somatic variants that are specific to the tumor DNA, any of which could be cancer-causing or -sustaining. TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%. Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}. To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large-scale genomics data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.

@@

Several factors limit the conclusions we may draw from this analysis. First, in this analysis we used repeatability between technical replicates (being in the WGS and WGA samples) as a measure of confidence in a putative SNV. This metric is potentially problematic for many reasons, including but not limited to: (i) cancer is highly heterogeneous, and so a legitimate somatic SNV might show up in one replicate but not another; (ii) if the DNA sample is degraded to some extent, due to surgery conditions or some other factor out of the hands of the sequencing center, the same errors may appear in both replicates; (iii) the SNV calling process may enrich for artifacts such as germ line variants, which are nearly indistinguishable from somatic SNVs in computational analyses; cross-referencing to gold standards such as dbSNP is the only way to identify germline SNPs; (iv) a putative mutation that is present in both samples may be a somatic mutation that arose before the tumor \citep{pre-tumor-muts}. Although repeatability across technical replicates cannot guarantee that a putative mutation is a true somatic mutation, it does increase the likelihood. Having an independent measure of confidence in SNV calls, even an imperfect one, can help us gauge the accuracy of other measures, specifically of different filtering approaches.
